---
title: "Getting Started with PITmodelR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PITmodelR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, eval=F, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 4,
  message = TRUE, warning = TRUE
)
```

# Overview

`PITmodelR` streamlines access to [PTAGIS](https://www.ptagis.org/) data through API endpoints and provides helper tools for PIT-tag data workflows, including project discovery, site metadata, tag observations, and survival/migration modeling.

This vignette shows how to:
- Install the package
- Configure API authentication
- Retrieve project codes and available years of data
- Explore MRR and interrogation site metadata
- Download mark-recapture-recovery (MRR) and observation PIT-tag data
- Retrieve tag event histories

> **Note:** Network calls and API-key examples are set to `eval = FALSE` to keep this vignette runnable on CRAN. Try them locally after setting your API key.

---

# Installation

The `PITmodelR` package can be installed from GitHub using the install function from the package `remotes`.

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("ryankinzer/PITmodelR")
```

As for any R package, it must be loaded into the session environment before the functions will operate.

```{r eval=FALSE}
library(PITmodelR)
```

---

# Authentication (API key)

Some PTAGIS endpoints require an API key. You can pass it directly to functions as a character string using the argument "PTAGIS_API_KEY" or set it once as a session environment variable.

```{r eval=FALSE}
Sys.setenv(PTAGIS_API_KEY = "YOUR-PTAGIS-API-KEY")
```

Another option is to add `PTAGIS_API_KEY` permanently to your `~/.Renviron` file for persistent use. The proper method and steps to complete this action are not included in this tutorial.  

---

# Project  Data

PTAGIS files are organized into projects with a three character code. In the past, these project codes were called "Coordinator IDs". A project code is appended to the name of every data file uploaded to PTAGIS. The function `get_project_codes()` returns a vector of all project codes that have data associated with them in PTAGIS.

```{r eval=FALSE}
codes <- get_project_codes()
```

Package users may need to check the number of years a specific project has submitted PIT-tag data. This task is easily completed with `get_project_years()`. 

```{r eval=FALSE}
yrs <- get_project_years(code = "CDR")
```


Individual mark-recapture-recovery (MRR) files submitted for a given project and year can be obtained with `get_mrr_files()`. The function returns a small data frame containing metadata about the file uploads. 

```{r eval=FALSE}
files <- get_mrr_files(code = "CDR", year = 2024)
files$name
```

After we know which files are available we can then download all uploaded data submitted with the file using `get_file_data()`. The function returns a list with four objects; session, events, session_pdv_fields, and detail_pdv_fields. The list can then be collapsed into a simple data frame using `flatten_mrr_file()` for easier manipulation. 

```{r eval=FALSE}
mrr_file <- get_file_data(files$name[1])
mrr_file$session
mrr_file$events
mrr_file$session_pdv_fields
mrr_file$detail_pdv_fields

df <- flatten_mrr_file(mrr_file)
```
Or, you can download data from multiple MRR files in a single step using the batch function `get_batch_file_data()`. Because uploaded MRR files may not have standardized PDV fields, the function also contains arguments for how the output is returned based on label consistency. The function returns a list with three objects: "files", "combined", and "issues". The "files" object is a nested list with individual objects for each file name supplied to the function. The second object, "combined", is a single data frame with the data flattened across all the files. The final object, "issues", displays discrepancies between user defined field labels (SPDV and PDV).

> **Note:** User defined fields may contain dates, numbers, or character strings, as such, all data in those fields are returned as a character string. Additional data formatting may be necessary if user defined fields require a specific format.

```{r eval=FALSE, message=FALSE}
filenames <- files$name[grepl('SCT', files$name)]

# Batch workflow:
all_data <- get_batch_file_data(
  filenames,
  check_labels = "warn",         # checks if user defined fields have the same labels, can use "error" to stop on mismatches
  keep_code_cols = FALSE,         # should we keep PDV/SPDV code columns and the user defined labels
  label_conflict = "suffix",     # behavior for file label column collisions
  use_codes_on_conflict = TRUE   # prefer consistent code columns across files
)

# list of all individual files
names(all_data$files)

# look at data for a single file
all_data$files$`CDR-2024-101-JCT.xml`

# combined session and event data into a single dataset
df <- all_data$combined

# label mismatch issues
all_data$issues
```
---
# Survival Studies

Survival studies begin with a list of tags that represent the population of interest and will later be recaptured or observed. The above functions allow us to download all the data associated with a MRR file (in this case a mark or tagging file) uploaded to PTAGIS. To create a list of tags suitable for a survival study, we first need to clean the MRR data and filter for tags that best represent our group of interest. This can easily be completed with functions from the `tidyverse` packages.


## Mark List

```{r eval=FALSE}
library(tidyverse)

mark_group <- df %>%
  filter(species_run_rear_type == '12W',
         migration_year == 2025,
         between(release_date, ymd(20240901), ymd(20241231)),
         release_site == 'SECTRP',
         pittag != "..........",
         !grepl('RE', text_comments),
         !grepl('Y', conditional_comments)
  )
```

---

## Pit-Tag Histories


```{r eval=FALSE}
pittag <- mark_group$pittag[1]
tag_history <- get_tag_history(tag_code = pittag)
```

```{r eval=FALSE}
pittags <- mark_group$pittag
tag_history <- get_batch_tag_histories(tag_codes = pittags)
```

## Recapture Observations

We first need to define our survival reaches by identify locations of potential recaptures. We can start by looking at a table with all the locations that the tag group was observed.

```{r eval=FALSE}
table(tag_history$event_type)
table(tag_history$site_code)
```

All the locations look good with the exception of "CWR".  To build my capture history I will first remove "CWR" and then select the sites that I need for survivals, SECTRP, ZEN, SFG, and Lower Granite Dam (GRJ and GRS), all other locations are downstream of Lower Granite (LGR) and can be used to calculate a detection efficiency at LGR. 

```{r eval=FALSE}

locs <- c("SECTRP","ZEN","SFG","LGR","Down")

locs <- list(
  SECTRP = "SECTRP",
  ZEN    = "ZEN",
  SFG    = "SFG",
  LGR    = c("GRJ","GRS"),
  Down   = c("LMN","MCN","BON", "B2J", "BCC", "GOJ", "ICH", "JDJ", "LMJ", "MCJ", "PD7", "PD8", "PDW", "TWX")
)

res <- build_mark_histories(
  tag_history  = tag_history,
  locs_def     = locs,
  site_col     = "site_code",
  tag_col      = "tag_code",
  time_col     = "event_time",  # optional but helps order ties/revisits
  enforce_order = TRUE,
  keep_unknown  = FALSE
)

ch_data <- res$ch_data      # tag_code, ch
ch_freq <- res$ch_freq      # ch, freq
res$dropped_summary         # any sites removed because not in `locs`
res$mapping                 # site -> occasion index
```


```{r eval=FALSE}
# Fit CJS model
library(marked)

fit <- fit_marked_cjs(res$ch_data,
                      phi_formula = ~ time,
                      p_formula   = ~ time,
                      conf_level  = 0.95)

# Peek tables
head(fit$phi)
head(fit$p)

# Show plots
if (inherits(fit$plots$phi, "ggplot")) {
  print(fit$plots$phi)
  print(fit$plots$p)
} else if (is.function(fit$plots$phi)) {
  fit$plots$phi()  # base plot
  fit$plots$p()
}

# Full model summary if needed
summary(fit$model)

fit$cum_phi         # cumulative survival with CIs
fit$covariance_mode # "full" if covariance used, otherwise "independence_fallback"

# Plot
if (inherits(fit$plots$cum_phi, "ggplot")) {
  print(fit$plots$cum_phi)
} else {
  fit$plots$cum_phi()
}

```

```{r eval=FALSE}
proc <- process.data(res$ch_data, model = "CJS")
ddl  <- make.design.data(proc)

Phi.time <- list(formula = ~ time)
p.time   <- list(formula = ~ time)

mod <- crm(proc, ddl, model.parameters = list(Phi = Phi.time, p = p.time))
summary(mod)

mod$results
mod$results$reals$Phi

est_preds = predict(mod) %>% 
      map(.f = as_tibble)
```


---

# Interrogation Sites

## List interrogation sites (metadata)

```{r eval=FALSE}
sites_active <- list_interrogation_sites()
sites_all <- list_interrogation_sites(active_only = FALSE)
sites_min <- list_interrogation_sites(fields = c("site_code", "site_name", "status"))
```

## Get metadata for a single site

```{r eval=FALSE}
md <- get_site_metadata("LGR")
md
```

---

# Observations

## Get site observations (with auto-pagination)

```{r eval=FALSE}
obs1 <- get_site_observations(site_code = "LGR")
obs2 <- get_site_observations(site_code = "LGR", year = 2023)
obs_page1 <- get_site_observations(site_code = "LGR", page = 1, page_size = 1000, all_pages = FALSE)
obs_small <- get_site_observations(
  site_code = "LGR", year = 2023,
  fields = c("event_time", "tag_code", "antenna", "site_code")
)
```

---

# Tag Event History

## Get observations for a single PIT tag

```{r eval=FALSE}
tag_events <- get_tag_history(tag_code = "384.1B79726A98")
tag_events_min <- get_tag_history(
  tag_code = "384.1B79726A98",
  fields = c("event_time", "site_code", "antenna", "direction")
)
```

---

# Error handling & best practices

- Functions use a resilient HTTP layer with retries and helpful error messages.
- When nothing is returned, functions **warn** (instead of stopping) and return an empty tibble or vector.
- Common transformations:
  - Names normalized to `snake_case`
  - Returns standardized to **tibble** for downstream use
- For long queries or production code, consider:
  - Caching responses (e.g., `memoise` package)
  - Rate limiting / backoff on large downloads
  - Filtering by time windows when available

---

# Session info

```{r}
sessionInfo()
```

---

# Appendix: Troubleshooting

- **HTTP 4xx** errors: verify your API key, parameters, and endpoint paths.
- **Empty results**: confirm `site_code`, `tag_code`, date/year filters, and that the site/tag actually has records.
- **Name mismatches**: after `snake_case` normalization, field names may differ slightly from raw API (e.g., `SiteCode` → `site_code`).
